Describing the scale and composition of calls for police service: a replication and extension using open data

### 2022-02-18 paper.Rmd
inserted 'to' in first sentence
small spelling error
paper complete

---
title: 'Describing the scale and composition of calls for police service: a replication and extension using open data'
author: Samuel Langton\dag, Stijn Ruiter\dag\*, Tim Verlaan\dag
bibliography: refs.bib
output:
  pdf_document: 
    number_sections: true
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
csl: apa.csl
abstract: "This paper describes the scale and composition of emergency demand for police services in Detroit, United States. The contribution is made in replication and extension of analyses reported by Ratcliffe (2021) and Lum et al. (2021). Findings indicate that police spend a considerable proportion of time performing a social service function. Just 51% of total deployed time responding to 911 calls is consumed by crime incidents. The remainder is spent on community (5%), health (7%), proactive (4%), quality of life (16%), and traffic (15%) duties. A small number of calls consume a disproportionately large amount of police officer time. Emergency demand is concentrated in time and space, and can differ between types of demand. Findings further highlight the potential implications of radically reforming police forces in the United States. The data and code used here are openly available for reproduction, re-use and scrutiny."
---

@@

Here, we propose an extension of @ratcliffe2021policing and @lum2021can. This extension has two principal components. First, we replicate the main analyses of these existing investigations into the scale and composition of police demand using open study materials. To date, investigations have been conducted on data which are not publicly available and without the materials (e.g., code) to reproduce the results. Open science practices have been identified as fundamental to the aims of evidence-based policing and vital in ensuring that in the long-term, real-life decisions are based on a concrete and replicable pool of evidence [@openedp2019]. In this case, the necessity for a fully open investigation into the scale and composition of police demand is particularly important given the public nature of the debate around police reform and the considerable implications of redefining the role of the police without transparent empirical investigation. It is reasonable for police practitioners and the public to expect a thorough, open and replicable investigation into police demand before academia concludes that the issue is resolved – certainly before findings are used to change or introduce police practice or public policy. Especially since classifying police demand into these different categories is no straightforward or facile task. The difference in categories in Ratcliffe's and Lum's paper attest to this and showcase why they should not have to be taken at face-value. Open materials have the additional benefit of being available for re-use in other jurisdictions, helping build a reliable and comparable evidence-base around the scale and composition of police demand across different contexts.

@@

The 'time spent' measure used was the _time on scene_ (i.e., excluding the response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported as resolved. In this way, the calculation did not account for the additional time consumed by multi-officer responses (e.g., back-up). This was justified for a number of reasons, among them, avoiding the over-inflation of time estimates due to officers 'flocking' to non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

Summary statistics describe the overall composition of demand according to the six demand type classifications. This includes call frequency counts, a proportional breakdown of call counts, and a proportional breakdown of police time. Both _total deployed time_ and _time on scene_ measures of police resource are reported. The proportional breakdown of _time on scene_ is visualized using a treemap graphic for each incident type nested within the demand classifications. The equivalent graphic for _total deployed time_ is available in the corresponding GitHub repository.

A series of descriptive statistics summarize the time on scene consumed by each demand type. For comprehensiveness and transparency, we also report the mean, median, standard deviation, and range in the Appendix, next to a histogram of the underlying distribution in minutes. This sheds on light on how long a 'typical' call might take while noting a probable skew in the distribution. Such a skew would suggest that averages of time [e.g., @lum2021can] or aggregate measures of time [e.g., @ratcliffe2021policing] in isolation are insufficient and potentially mask important underlying variation, namely, that a small number of high-demand calls drag up the mean and inflate aggregate measures of time. To investigate this distribution, we plot Lorenz curves for each demand classification. Lorenz curves plot the cumulative distribution of a variable (in this case, time on scene) against the cumulative distribution of a unit (in this case, calls) [@bernasco2017more]. By way of example, we report on these figures according to arbitrary thresholds of 25% and 50% of time spent. This sheds light on the degree to which a disproportionately large amount of officer time might be consumed by just a small number of calls.

@@

The temporal patterning of the six demand type classifications are visualized using a heatmap of mean call counts, aggregated by day of the week, and hour of the day. To summarize the spatial patterning, call location coordinates are aggregated to synthetic grid cells using spatial joins. Grid cells are defined as 1000 ft$^2$ (305m$^2$) -- a size deemed appropriate based on a balance between capturing localized variation and interpretable visualizations. 2% of incidents are excluded from the spatial visualizations due to incomplete geographic information. To accurately convey the resource deployed to each area during the year, maps visualize the aggregate _time on scene_ (in hours) within each grid cell over the course of the year, for each demand type classification. The equivalent raw count maps are available in the Appendix. In this way, we visualize the spatio-temporal patterning of demand in a manner comparable to @ratcliffe2021policing, while adding a comparison across different demand classifications and two different measures of demand.

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consume 51% of time on scene throughout the year. The remainder of time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%). The differences in the proportional breakdown between time on scene and total deployment time, which includes response time, are marginal. This indicates that the response time plays only a minor role determining the scale and composition of time consumed.

@@

To further quantify the resources consumed, Table 2 summarizes the time spent on the scene for each demand classification in minutes. Histograms of the underlying distributions are reported in the Appendix. Table 2 demonstrates a number of things. First, there is a modest amount of variation in the typical time at scene of different types of emergency call. For instance, the mean time spent on the scene of traffic incidents (55 minutes) and crime (51 minutes) is considerably more that quality of life issues (31 minutes) or proactive activity (28 minutes). These differences will be a result of a combination of factors, including the complexity and severity of incidents. Second, there is considerable variation in the time consumed by incidents _within_ each demand type. While many calls are resolved in minutes, a handful consume hours of officer time on the scene, resulting in a highly skewed distribution. This highlights how mean or aggregate summary statistics of police time as a measure of demand can mask (important) underlying variation.

@@

In replication and extension of the main findings reported by @ratcliffe2021policing and @lum2021can this contribution has described the scale and composition of calls for police service in Detroit, US. Broadly speaking, the results presented here provide robustness to those reported by @ratcliffe2021policing and @lum2021can. There is now substantial empirical evidence to demonstrate that, in multiple jurisdictions, the public relies on the police for both social _and_ crime-fighting functions. A reduction in the capacity of the police to meet this emergency demand, without first providing appropriate, capable and accessible replacement emergency services, could have considerable implications for public safety and well-being. This contribution provides the first such demonstration using open data and open code, opening up analyses for scrutiny, re-use, and extension by the public, academics and/or police practitioners. We encourage usage of the materials provided with this study to replicate analyses across other jurisdictions. In doing so, we hope to see a consistent and reliable evidence-base across different contexts for the purposes of comparison and generalization.

In this study, calls were categorized according the six classes used by @ratcliffe2021policing. The overall composition of police demand in terms of call volume and proportional time spent are remarkably similar. A considerable volume of calls for service, and in turn, a considerable proportion of deployed police time, relate to incident types which do not fall under the traditional 'crime-fighting' role of the police. Instead, the police fulfill a variety of other social service functions, including the resolution of public (mental) health incidents and resolving community issues. That said, descriptive statistics which drill down into the underlying distribution of officer time consumed by each demand type provide additional insight which would otherwise be overlooked. First, it is clear that there is considerable variation in the amount of time consumed both between and within demand classifications. On a methodological point, this highlights how average (e.g., mean times) or aggregate (e.g., sum times) might mask underlying variation. Second, and relatedly, these extensions demonstrate that a disproportionately large amount of police time is spent on a small number of emergency calls. 50% of total time on scene in consumed by just 15% of calls. High-demand calls will likely be complex and/or life-threatening incidents or calls which require officers to secure a scene for extended periods of time. A targeted examination of these calls might shed light on the extent to which such high-intensity demands could be reduced or better resolved through or in concert with other public services.

On this point, this study raises useful discussion points around the different measures of officer time. First, it is clear that findings on the scale and composition of police demand are only marginally sensitive to whether measures of officer time include the response time -- defined as the time from dispatch to when the first officer arrives on-scene. The overall findings are almost identical between the two measures. Here, we would highlight that our response time measure did not include dispatch time, but rather, travel time in isolation. Including dispatch time would increase the gap between the two measures: low-priority calls will be held in the dispatch queue for longer, inflating the overall response time, while high-priority calls will be dispatched immediately. Since certain calls will have higher priority responses (e.g., violent crime) this gap is likely to differ between demand classifications. We excluded dispatch time on the basis that it does not capture deployment, but a specific investigation into its impact would be straightforward using the open materials provided.

Second, and relatedly, we would reiterate that the 'time on scene' measure reported here, in alignment with @lum2021can, refers to the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. This differs to that of @ratcliffe2021policing which accounted for the number of officers attending the scene. @lum2021can note that including back-up officers in the time calculation might introduce complications. For instance, non-serious incidents might receive back-up unnecessarily, particularly during periods of low-demand, which would inflate the time consumed and overstate the resources required to resolve such an incident. Although the overall composition of police demand is comparable between Philadelphia [@ratcliffe2021policing] and Detroit, the true impact of these measures remains unclear. We propose that future research investigates these distinctions. This would require CAD (or equivalent) data, timestamped at each stage of the response (dispatch, scene arrival, scene departure), for individual police units nested within calls. This level of granularity would permit the calculation and comparison of multiple measures of officer time, as well as an investigation into the extent to which back-ups and/or 'flocking' might influence such measures.

>>>>

---
title: 'Describing the scale and composition of calls for police service: a replication and extension using open data'
author: Samuel Langton\dag, Stijn Ruiter\dag\*, Tim Verlaan\dag
bibliography: refs.bib
output:
  pdf_document: 
    number_sections: true
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
csl: apa.csl
abstract: "This paper describes the scale and composition of emergency demand for police services in Detroit, United States. The contribution is made in replication and extension of analyses reported by Ratcliffe (2021) and Lum et al. (2021). Findings indicate that police spend a considerable proportion of time performing a social service function. Just 51% of total deployed time responding to 911 calls is consumed by crime incidents. The remainder is spent on quality of life (16%), traffic (15%), health (7%), community (5%) and proactive (4%) duties. A small number of calls consume a disproportionately large amount of police officer time. Emergency demand is concentrated in time and space, and can differ between types of demand. Findings further highlight the potential implications of radically reforming police forces in the United States. The data and code used here are openly available for reproduction, re-use and scrutiny."
---

@@

Here, we propose an extension of @ratcliffe2021policing and @lum2021can. This extension has two principal components. First, we replicate the main analyses of these existing investigations into the scale and composition of police demand using open study materials. To date, investigations have been conducted on data which are not publicly available and without the materials (e.g., code) to reproduce the results. Open science practices have been identified as fundamental to the aims of evidence-based policing and vital in ensuring that in the long-term, real-life decisions are based on a concrete and replicable pool of evidence [@openedp2019]. In this case, the necessity for a fully open investigation into the scale and composition of police demand is particularly important given the public nature of the debate around police reform and the considerable implications of redefining the role of the police without transparent empirical investigation. It is reasonable for police practitioners and the public to expect a thorough, open and replicable investigation into police demand before academia concludes that the issue is resolved – certainly before findings are used to change or introduce police practice or public policy. Open materials have the additional benefit of being available for re-use in other jurisdictions, helping build a reliable and comparable evidence-base around the scale and composition of police demand across different contexts.

@@

The 'time spent' measure used was the _time on scene_ (i.e., excluding the response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported as resolved. In this way, the calculation did not account for the additional time consumed by multi-officer responses (e.g., back-up). This was justified for a number of reasons, among them, avoiding the over-inflation of time estimates due to multiple officers attending non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

Summary statistics describe the overall composition of demand according to the six demand type classifications. This includes call frequency counts, a proportional breakdown of call counts, and a proportional breakdown of police time. Both _total deployed time_ (response time + time on scene) and _time on scene_ measures of police resource are reported. The proportional breakdown of _time on scene_ is visualized using a treemap graphic for each incident type nested within the demand classifications. The equivalent graphic for _total deployed time_ is available in the corresponding GitHub repository.

A series of descriptive statistics summarize the time on scene consumed by each demand type. For comprehensiveness and transparency we report the mean, median, standard deviation and range, and in the Appendix, a histogram of the underlying distribution. This sheds on light on how long a 'typical' call might take while noting a probable skew in the distribution. Such a skew would suggest that averages of time [e.g., @lum2021can] or aggregate measures of time [e.g., @ratcliffe2021policing] in isolation are insufficient and potentially mask important underlying variation, namely, that a small number of high-demand calls drag up the mean and inflate aggregate measures of time. To investigate this distribution, we plot Lorenz curves for each demand classification. Lorenz curves plot the cumulative distribution of a variable (in this case, time on scene) against the cumulative distribution of a unit (in this case, calls) [@bernasco2017more]. By way of example, we report on these figures according to arbitrary thresholds of 25% and 50% of time spent. This sheds light on the degree to which a disproportionately large amount of officer time might be consumed by just a small number of calls.

@@

The temporal patterning of the six demand type classifications are visualized using a heatmap of mean call counts, aggregated by day of the week, and hour of the day. To summarize the spatial patterning, call location coordinates are aggregated to synthetic grid cells using spatial joins. Grid cells are defined as 1000 ft$^2$ (305m$^2$) -- a size deemed appropriate based on a balance between capturing localized variation and interpretable visualizations. 2% of incidents are excluded from the spatial visualizations due to incomplete geographic information. To accurately convey the resource deployed to each area during the year, maps visualize the aggregate _time on scene_ (in hours) within each grid cell over the course of the year for each demand type classification. The equivalent raw count maps are available in the Appendix. In this way, we visualize the spatio-temporal patterning of demand in a manner comparable to @ratcliffe2021policing, while adding a comparison across different demand classifications and two different measures of demand.

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consume 51% of time on scene throughout the year. The remainder of time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%). The differences in the proportional breakdown between time on scene and total deployment time, which includes response time, are marginal. This indicates that the response time--as measured by travel time--plays only a minor role determining the scale and composition of time consumed.

@@

To further quantify the resources consumed, Table 2 summarizes the time spent on the scene for each demand classification in minutes. Histograms of the underlying distributions are reported in the Appendix. Table 2 demonstrates a number of things. First, there is a modest amount of variation in the typical time at scene of different types of emergency call. For instance, the mean time spent on the scene of traffic incidents (55 minutes) and crime (51 minutes) is considerably more that quality of life issues (31 minutes) or proactive activity (28 minutes). These differences will be a result of a combination of factors, including the complexity and severity of incidents. Second, there is considerable variation in the time consumed by incidents _within_ each demand type. While many calls are resolved in minutes, a handful consume hours of officer time on the scene, resulting in a highly skewed distribution. This also highlights how mean or aggregate summary statistics of police time as a measure of demand can mask (important) underlying variation.

@@

In replication and extension of the main findings reported by @ratcliffe2021policing and @lum2021can this contribution has described the scale and composition of calls for police service in Detroit, US. Broadly speaking, the results presented here provide robustness to those reported by @ratcliffe2021policing and @lum2021can. There is substantial empirical evidence to demonstrate that, in multiple jurisdictions, the public relies on the police for both social _and_ crime-fighting functions. A reduction in the capacity of the police to meet this emergency demand, without first providing appropriate, capable and accessible replacement emergency services, could have considerable implications for public safety and well-being. This contribution provides the first such demonstration using open data and open code, opening up analyses for scrutiny, re-use, and extension by the public, academics and/or police practitioners. We encourage usage of the materials provided with this study to replicate analyses across other jurisdictions. In doing so, we hope to see a consistent and reliable evidence-base across different contexts for the purposes of comparison and generalization.

In this study, calls were categorized according to the six classes used by @ratcliffe2021policing. The overall composition of police demand in terms of call volume and proportional time spent are remarkably similar. A considerable volume of calls for service, and in turn, a considerable proportion of deployed police time, relate to incident types which do not fall under the traditional 'crime-fighting' role of the police. Instead, the police fulfill a variety of other social service functions, including the resolution of public (mental) health incidents and resolving community issues. That said, descriptive statistics which drill down into the underlying distribution of officer time consumed by each demand type provide additional insight which would otherwise be overlooked. First, it is clear that there is considerable variation in the amount of time consumed both between and within demand classifications. On a methodological point, this highlights how average (e.g., mean times) or aggregate (e.g., sum times) might mask underlying variation. Second, and relatedly, these extensions demonstrate that a disproportionately large amount of police time is spent on a small number of emergency calls. 50% of total time on scene in consumed by just 15% of calls. High-demand calls will likely be complex and/or life-threatening incidents or calls which require officers to secure a scene for extended periods of time. A targeted examination of these calls might shed light on the extent to which such high-intensity demands could be reduced or better resolved through or in concert with other public services.

On this point, this study raises useful discussion points around the different measures of officer time. First, it is clear that findings on the scale and composition of police demand are only marginally sensitive to whether measures of officer time include the response time -- defined as the time from dispatch to when the first officer arrives on-scene. The overall findings are almost identical between the two measures. Here, we would highlight that our response time measure did not include dispatch time, but rather, travel time in isolation. Including dispatch time would increase the gap between the two measures: low-priority calls will be held in the dispatch queue for longer, inflating the overall response time, while high-priority calls will be dispatched rapidly. Since certain calls will have higher priority responses (e.g., violent crime) this gap is likely to differ between demand classifications. We excluded dispatch time on the basis that it does not capture deployment, but a specific investigation into its impact would be straightforward using the open materials provided.

Second, and relatedly, we would reiterate that the 'time on scene' measure reported here, in alignment with @lum2021can, refers to the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. This differs to that of @ratcliffe2021policing which accounted for the number of officers attending the scene. @lum2021can note that including back-up officers in the time calculation might introduce complications. For instance, non-serious incidents might receive back-up unnecessarily, particularly during periods of low-demand, which would inflate the time consumed and overstate the resources required to resolve such an incident. Although the overall composition of police demand is comparable between Philadelphia [@ratcliffe2021policing] and Detroit, the true impact of these measures remains unclear. We propose that future research investigates these distinctions. This would require CAD (or equivalent) data, timestamped at each stage of the response (dispatch, scene arrival, scene departure), for individual police units nested within calls. This level of granularity would permit the calculation and comparison of multiple measures of officer time, as well as an investigation into the extent to which back-ups and/or 'flocking' might influence such measures. In sum, further investigations into different measures of 'police time' and the corresponding impact on our assessment of the scale and composition of police demand are necessary before we can claim that the evidence-base is sufficient.
